```{r setup, include=FALSE}
library(flowCore)
library(tidyverse)
library(flowSpectrum)
library(ggcyto)
library(pheatmap)
library(matrixStats)
library(corrplot)
library(data.table)
set.seed(233)
library(reticulate)
```

```{python}
# Load packages
import numpy as np
from sklearn import linear_model
from sklearn.model_selection import train_test_split, GridSearchCV
from sklearn.metrics import mean_squared_error
import pandas as pd
import fcsparser
np.random.seed(233)
```

If you really hate those negative values in unmixed data, NNLS might be a good option for you. Compared to OLS, it adds a non-negative constraint to the abundances. So the negative values get assigned 0 instead. I personally don't think this one is that good in practice, cause it literally makes the values near 0 a very thin line. It's harder to identify the -ve population and where the gates are. At least in my data. But you are welcomed to give it a go.

Before start, build the spectral matrix then convert your spectral matrix from R into a .csv to load in as a python object.

-   You can also use packages like rds2py to convert data between these 2 languages.

```{r}
# load reference controls
file.dir <-"./ref controls"
files <- dir("./ref controls", pattern = "fcs$", full.names=T)

# read in fcs files
dat <- lapply(files, read.FCS)
names(dat) <-gsub("./ref controls/","",files)

p1 <- lapply(seq(2,length(dat), by = 2), function(i){spectralplot(dat[[i]], normalize = TRUE, guessPop = TRUE, unstained = dat[[i-1]])$data})

#Name the list of new dfs
 names(p1) <- gsub("./ref control/", "",names(dat[seq(2,length(dat),2)]))
 names(p1) <- gsub("_pos.fcs", "",names(dat[seq(2,length(dat),2)]))
 print(as.matrix(names(p1)))

### If using flowjo exported files: switch to this exact order of channels.

aurora_channels <- c('UV1-A',  'UV2-A',  'UV3-A',  'UV4-A',  'UV5-A',  'UV6-A',  'UV7-A',  'UV8-A',  'UV9-A',  'UV10-A','UV11-A', 'UV12-A', 'UV13-A', 'UV14-A', 'UV15-A', 'UV16-A', 'V1-A',   'V2-A',  'V3-A',   'V4-A',   'V5-A',   'V6-A',   'V7-A',   'V8-A',   'V9-A',   'V10-A',  'V11-A',  'V12-A',  'V13-A',  'V14-A',  'V15-A',  'V16-A',  'B1-A',   'B2-A',   'B3-A',   'B4-A',  'B5-A',   'B6-A',   'B7-A',   'B8-A',   'B9-A',   'B10-A',  'B11-A',  'B12-A',  'B13-A',  'B14-A',  'YG1-A',  'YG2-A',  'YG3-A',  'YG4-A',  'YG5-A',  'YG6-A',  'YG7-A',  'YG8-A', 'YG9-A',  'YG10-A', 'R1-A',   'R2-A',   'R3-A',   'R4-A',   'R5-A',   'R6-A',   'R7-A',   'R8-A'  )

# You can also save the aurora_channels.rds for future easy use.

# Reorder the channels
for (i in 1:length(p1)){
  p1[[i]] <- p1[[i]][match(aurora_channels, p1[[i]]$`rownames(df)`),]
}

#### fetch the channel names directly if you export data from cytek software
 cvars<-gsub("\ ","",apply(expand.grid(c("UV", 'V', 'B', "YG", "R"),1:16,"-A") %>% 
                            mutate(Var1=factor(Var1,levels=unique(Var1))) %>% 
                            arrange(Var1,Var2),1,function(X) paste0(X,collapse="")))

# add a sample column for plotting
  for(i in 1:length(p1)){p1[[i]]$sample = names(p1)[i]}

# combine the df into a big df
df <- df <- dplyr::bind_rows(c(p1))

# Add the AF parameter in
dir <- "./af"
files <- dir(dir, pattern = "fcs$", full.names=T)
as.matrix(files)

# read in fcs files
dat <- lapply(files, read.FCS)

# tidy up the file names as you wish, if you need to
#names(dat) <-gsub("./af/export_","",files)
#names(dat) <-gsub("_2026.*","",names(dat))

# Extract normalised spectra
af <- lapply(1:length(dat), function(i){spectralplot(dat[[i]], normalize = TRUE)$data})
names(af) <- names(dat)

# Reorder the channels
af<- af[match(aurora_channels, af$`rownames(df)`),]

### construct the unmixing matrix with AF
# add AF to the control set
spectra_matrix <-as.data.frame(spectra_matrix)
spectra_matrix$AF <- af$`medians.max.medians.`

write.csv(spactra_matrix,"spectral_matrix.csv")
```

Load data.

```{python}
# read data
path = "path/to/full-stained raw data.fcs"

# parse both metadata and data, get the expression values into a data frame
meta,data = fcsparser.parse(path, meta_data_only=False, reformat_meta=True)

# have a look
print(type(data))
data

# read in the spectral matrix
M = pd.read_csv("./spectral_matrix.csv")
print(type(M))

# take only the fluorescence channels from the raw data
Y = data.filter(regex="(UV|V|B|YG|R)(\d+)")
print(Y.columns)

# store the other half of the data
C = data.drop(columns=Y.columns)

# transpose to have channels as row names
YT = Y.T

# final checks
print(YT.shape)
print(M.shape)
print(type(M))
print(type(YT))
print(M.columns)
```

```{python}
# build the model
# train-test split
x_train, x_test, y_train, y_test = train_test_split(M, YT, test_size=0.25,
                                                    random_state=42)
# fit the model
m1 = linear_model.LinearRegression(positive=True)
nnls = m1.fit(x_train, y_train)

# Look at the residuals
y_pred_nnls = nnls.predict(x_test)
print(mean_squared_error(y_test, y_pred_nnls))

# Get the coefficients
print(nnls.coef_.shape)

# generate a df
feature_names = M.columns

A = pd.DataFrame(
    nnls.coef_,
    columns=feature_names
)
print(A)

# stick A back to the non-fluorescence channels
unmixed=pd.concat([A,C],axis=1)

# Write the .csv
unmixed.to_csv('nnls_unmixed_cum500.csv', index=False)

```

Load back the unmixed df in R env, and write to FCS.

```{r}
# process and write fcs
a <- read.csv("nnls_unmixed_cum500.csv")
a<- as.matrix(a)
a_ff<-flowFrame(a)
write.FCS(a_ff, "nnls_unmixed_cum500wAF.fcs")
saveRDS(a, "20260426_nnls_unmixed_cum500wAF.rds")
rm(ols, M, a, a_ff)
gc()
```

Note: I put python and R codes interchangbly for an all-in-one workflow. But if you have VS Code and are comfortable at using it, I still suggest you to use that for running python codes then jump back to Rstudio to run all those data format conversions. Python kernel has to go through reticulate in quarto documents, so it's slower to run here. VS Code also has autofilling for python codes if you like it.
