---
title: "Unmixing in R"
author: "Yuchen Li"
date: "2026-Apr"
output: 
  html_document:
    code_folding: "hide"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(flowCore)
library(flowSpectrum)
library(ggcyto)
library(pheatmap)
library(matrixStats)
library(corrplot)
library(data.table)
set.seed(233)
dir.create("./rds")
dir.create("./plots")
```

## Load single-stained controls

Use pre-gated control files from flowjo or other software of your prefer before this.\ 

They should be stored in a folder names ref controls. And with _pos and _neg portions labeled, for the next chunk to run smoothly.\ 

Load pre-gated fcs files for reference controls.

```{r, message=FALSE, warning=FALSE}
file.dir <-"./ref controls"
files <- dir("./ref controls", pattern = "fcs$", full.names=T)

# read in fcs files
dat <- lapply(files, read.FCS)
names(dat) <-gsub("./ref controls/","",files)
```

Generate the data frame of reference control spectral profiles.

```{r, message=FALSE, warning=FALSE, echo=FALSE}
p1 <- lapply(seq(2,length(dat), by = 2), function(i){spectralplot(dat[[i]], normalize = TRUE, guessPop = TRUE, unstained = dat[[i-1]])$data})

#Name the list of new dfs
 names(p1) <- gsub("./ref control/", "",names(dat[seq(2,length(dat),2)]))
 names(p1) <- gsub("_pos.fcs", "",names(dat[seq(2,length(dat),2)]))
 print(as.matrix(names(p1)))

### If using flowjo exported files: switch to this exact order of channels.

aurora_channels <- c('UV1-A',  'UV2-A',  'UV3-A',  'UV4-A',  'UV5-A',  'UV6-A',  'UV7-A',  'UV8-A',  'UV9-A',  'UV10-A','UV11-A', 'UV12-A', 'UV13-A', 'UV14-A', 'UV15-A', 'UV16-A', 'V1-A',   'V2-A',  'V3-A',   'V4-A',   'V5-A',   'V6-A',   'V7-A',   'V8-A',   'V9-A',   'V10-A',  'V11-A',  'V12-A',  'V13-A',  'V14-A',  'V15-A',  'V16-A',  'B1-A',   'B2-A',   'B3-A',   'B4-A',  'B5-A',   'B6-A',   'B7-A',   'B8-A',   'B9-A',   'B10-A',  'B11-A',  'B12-A',  'B13-A',  'B14-A',  'YG1-A',  'YG2-A',  'YG3-A',  'YG4-A',  'YG5-A',  'YG6-A',  'YG7-A',  'YG8-A', 'YG9-A',  'YG10-A', 'R1-A',   'R2-A',   'R3-A',   'R4-A',   'R5-A',   'R6-A',   'R7-A',   'R8-A'  )

# You can also save the aurora_channels.rds for future easy use.

# Reorder the channels
for (i in 1:length(p1)){
  p1[[i]] <- p1[[i]][match(aurora_channels, p1[[i]]$`rownames(df)`),]
}

#### fetch the channel names directly if you export data from cytek software
 cvars<-gsub("\ ","",apply(expand.grid(c("UV", 'V', 'B', "YG", "R"),1:16,"-A") %>% 
                            mutate(Var1=factor(Var1,levels=unique(Var1))) %>% 
                            arrange(Var1,Var2),1,function(X) paste0(X,collapse="")))

# add a sample column for plotting
  for(i in 1:length(p1)){p1[[i]]$sample = names(p1)[i]}

# combine the df into a big df
df <- df <- dplyr::bind_rows(c(p1))
```

Visualise spectral profiles.

```{r}
# check the df structure is fine
# str(df)

# clean up the sample names if needed
#df$sample <- gsub(" ", "_", df$sample)

markers <- unique(df$sample)

# Visualise the spectral profiles
df %>% 
  mutate(Detector=factor(`rownames(df)`,levels=aurora_channels)) %>% 
    ggplot(aes(Detector,medians.max.medians.,group=sample,color=sample)) + 
      geom_line(linewidth = 0.7) + geom_point() + 
       guides(x=guide_axis(angle=45)) + theme_bw(base_size=8) + 
        theme(legend.text=element_text(size=5)) + 
        theme(legend.position = "right")+
        theme(legend.title = element_text(size=0.1))+
        ylab("")+
        xlab("")

#save the plot if you want
ggsave("./plots/ref control spectral profiles.png", width = 40, height = 10, units = c("cm"))
```

## Unmixing using OLS via SVD

Generate the spectral profile matrix and spectral unmixing matrix using SVD.

```{r, fig.height=6, fig.width=15}
### construct the spectra matrix
Mlist <- list()
for (i in 1:length(table(df$sample))){
    Mlist[[i]] <- as.matrix(p1[[i]]$medians.max.medians.)
}
spectra_matrix <- do.call(cbind, Mlist)
rm(Mlist)
colnames(spectra_matrix) <- names(p1)

#any(spectra_matrix<0) # subtracting the negative portion of the controls might cause some negative values

### construct the unmixing matrix

# SVD for M
svd_M <- svd(spectra_matrix)
pseudo_inver_M <- (svd_M$v) %*% diag(svd_M$d^(-1)) %*% t(svd_M$u)
rownames(pseudo_inver_M) <- colnames(spectra_matrix)
colnames(pseudo_inver_M) <-aurora_channels

# heatmap visualisation of the unmixing matrix
sample_name <- "Your sample name here"
pheatmap(pseudo_inver_M, 
         main = "Unmixing matrix by OLS",
         color = colorRampPalette(c("blue", "white", "red"))(100),
         cluster_rows = FALSE,
         cluster_cols = FALSE,
         angle_col = c("45"),
         filename = paste0("./plots/OLS_unmixing_matrix_", sample_name, ".png"),
       #  cellwidth = 5.5,
        width = 12,
        height = 4)

```

Load in raw sample data.

```{r}
fs_path <- "path/to/.fcs"
sample <- read.FCS(fs_path, transformation = "linearize-with-PnG-scaling")
print(sample@parameters@data$name[!sample@parameters@data$name %in% df$`rownames(df)`])
```

Obviously, 'Time',  "SSC-H"   "SSC-A"   "FSC-H"   "FSC-A" "SSC-B-H" "SSC-B-A" are not included in constructing the unmixing matrix.

```{r}
# take only the fluorescence channels to go into the unmixing
sample_fluo<-exprs(sample[,sample@parameters@data$name[sample@parameters@data$name %in% df$`rownames(df)`]])

as.matrix(sample_fluo)

dim(sample_fluo)

colnames(sample_fluo) <- NULL

# transpose it into the cells as columns vs channels as rows format
t(sample_fluo)

# Applying unmixing matrix
unmixed_matrix <- pseudo_inver_M %*% t(sample_fluo)

# Transpose to make the cells on rows again to match the original data
t_unmixed_matrix <- t(unmixed_matrix)
rm(unmixed_matrix, sample_fluo)

colnames(t_unmixed_matrix) <- gsub("_","_unmixed_", colnames(t_unmixed_matrix))

#Add these columns to the Time and FSC, SSC channels
unmixed_data <- cbind(exprs(sample[,sample@parameters@data$name[!sample@parameters@data$name %in% df$`rownames(df)`]]),
      t_unmixed_matrix)
rm(p1,dat)
```

Then we write the fcs file using flowCore to view it in conventional softwares as you like. Or you can stay in R and proceed with your visualisation and analysis.

```{r}
# write the fcs file to view in other UI softwares
ff <- flowFrame(unmixed_data)
write.FCS(ff, filename = paste0("ols_unmixed_", sample_name, ".fcs"))

# Save the unmixing results
OLS_unmixed <- list(
  ref_controls = df,
  Spectral_matrix = spectra_matrix,
  SVD = svd_M,
  Unmixing_matrix = as.matrix(pseudo_inver_M),
  Unmixed_fluo_channels = t_unmixed_matrix,
  Unmixed_sample = as.data.frame(unmixed_data),
  raw_sample_flowframe = sample,
  unmixed_sample_flowframe = ff
)
saveRDS(OLS_unmixed,paste0("./rds/OLS_unmixed_",sample_name, ".rds"))

# clean up temporary objects
rm(ff, sample, pseudo_inver_M, spectra_matrix, svd_M, t_unmixed_matrix, df)
```


## Calculate unmixing from the standard math notation (Novo, 2022)

This is based on the calculations from Novo et al. I would recommend that you give it a go and see how it looks using your data. The values in my data became less stable using this method compared to using SVD, and it became impossible for me to gate cells. However, if this method works well for you, you should stick with this one as this is the standard math model for unmixing.

```{r}
M <- spectra_matrix
# Calculate M^T
M <- as.data.table(M)
M_t <- transpose(M)
M <- as.matrix(M)
M_t <- as.matrix(M_t)
```

### Calculate the variance, co-variance matrix: (M^T M)^(-1)

```{r, fig.width=12, fig.height=10}
mtm <- (M_t %*% M)^(-1)
rownames(mtm) <- colnames(mtm)
```

### Calculate the closed-form solution for unmixed data

```{r}
# Load in raw data of full-stained sample
fs_path <- "path/to/.fcs"
sample <- read.FCS(fs_path, transformation = "linearize-with-PnG-scaling")
aurora_channels <- readRDS("./rds/aurora_channels.rds")
sample_fluo<-exprs(sample[,sample@parameters@data$name[sample@parameters@data$name %in% dat$ref_controls$`rownames(df)`]])

# quick check if you like
dim(sample_fluo)

# transpose the raw data into the cells as columns vs channels as rows format
sample_fluo<- as.data.table(sample_fluo)
sample_fluo<- transpose(sample_fluo)
sample_fluo <- as.matrix(sample_fluo)

dim(sample_fluo)
dim(M_t)
dim(mtm)

# Calculate the unmixed data
unmixed_fluo <- mtm %*% M_t %*% sample_fluo
```

### Write out the unmixed data

```{r}
# Transpose to make the cells on rows again to match the original data
t_unmixed_matrix <- t(unmixed_fluo)
colnames(t_unmixed_matrix) <- gsub("_","_unmixed_", colnames(t_unmixed_matrix))

#Add these columns to the Time and FSC, SSC channels
unmixed_data <- cbind(exprs(sample[,sample@parameters@data$name[!sample@parameters@data$name %in% dat$ref_controls$`rownames(df)`]]),
      t_unmixed_matrix)

# write the fcs file
ff <- flowFrame(unmixed_data)
sample_name <- "Your sample name"
write.FCS(ff, filename = paste0("ols_mtm_unmixed_", sample_name, ".fcs"))

# Save the results to rds if you want
```


## OLS unmixing with AF extraction

Unmixing with AF extraction extracts the spectrum of the matched unstained sample and model the total signals using fluorophores + an AF parameter.

Simply load in the cleaned-up usntained sample as well when you build the spectral matrix.

```{r}
dir <- "./af"
files <- dir(dir, pattern = "fcs$", full.names=T)
as.matrix(files)

# read in fcs files
dat <- lapply(files, read.FCS)

# tidy up the file names as you wish, if you need to
#names(dat) <-gsub("./af/export_","",files)
#names(dat) <-gsub("_2026.*","",names(dat))

# Extract normalised spectra
af <- lapply(1:length(dat), function(i){spectralplot(dat[[i]], normalize = TRUE)$data})
names(af) <- names(dat)

# Reorder the channels
af<- af[match(aurora_channels, af$`rownames(df)`),]

### construct the unmixing matrix with AF
# add AF to the control set
spectra_matrix <-as.data.frame(spectra_matrix)
spectra_matrix$AF <- af$`medians.max.medians.`
```

Then it's all the same below.

```{r}
# SVD for M
svd_M <- svd(spectra_matrix)
pseudo_inver_M <- (svd_M$v) %*% diag(svd_M$d^(-1)) %*% t(svd_M$u)
rownames(pseudo_inver_M) <- colnames(spectra_matrix)
colnames(pseudo_inver_M) <-aurora_channels

# heatmap visualisation of the unmixing matrix
sample_name <- "Your sample name here"
pheatmap(pseudo_inver_M, 
         main = "Unmixing matrix by OLS",
         color = colorRampPalette(c("blue", "white", "red"))(100),
         cluster_rows = FALSE,
         cluster_cols = FALSE,
         angle_col = c("45"),
         filename = paste0("./plots/OLS_unmixing_matrix_", sample_name, ".png"),
       #  cellwidth = 5.5,
        width = 12,
        height = 4)

```

Load in raw sample data.

```{r}
fs_path <- "path/to/.fcs"
sample <- read.FCS(fs_path, transformation = "linearize-with-PnG-scaling")
```


```{r}
# take only the fluorescence channels to go into the unmixing
sample_fluo<-exprs(sample[,sample@parameters@data$name[sample@parameters@data$name %in% af$`rownames(df)`]])

as.matrix(sample_fluo)

dim(sample_fluo)

colnames(sample_fluo) <- NULL

# transpose it into the cells as columns vs channels as rows format
t(sample_fluo)

# Applying unmixing matrix
unmixed_matrix <- pseudo_inver_M %*% t(sample_fluo)

# Transpose to make the cells on rows again to match the original data
t_unmixed_matrix <- t(unmixed_matrix)
rm(unmixed_matrix, sample_fluo)

colnames(t_unmixed_matrix) <- gsub("_","_unmixed_", colnames(t_unmixed_matrix))

#Add these columns to the Time and FSC, SSC channels
unmixed_data <- cbind(exprs(sample[,sample@parameters@data$name[!sample@parameters@data$name %in% af$`rownames(df)`]]),
      t_unmixed_matrix)
rm(dat)
```

Then we write the fcs file.

```{r}
# write the fcs file to view in other UI softwares
ff <- flowFrame(unmixed_data)
write.FCS(ff, filename = paste0("ols_unmixed_", sample_name, ".fcs"))

# Save the unmixing results
OLS_unmixedwAF <- list(
  Spectral_matrix = spectra_matrix,
  SVD = svd_M,
  Unmixing_matrix = as.matrix(pseudo_inver_M),
  Unmixed_fluo_channels = t_unmixed_matrix,
  Unmixed_sample = as.data.frame(unmixed_data),
  raw_sample_flowframe = sample,
  unmixed_sample_flowframe = ff
)
saveRDS(OLS_unmixed,paste0("./rds/OLS_unmixed_wAF_",sample_name, ".rds"))

# clean up temporary objects
rm(ff, sample, pseudo_inver_M, spectra_matrix, svd_M, t_unmixed_matrix, df)
```

Note: This is the most straightforward AF extraction method. From my experience and on consultation of Dr Novo, adding in the very matching AF background (this including washing and perming your unstained sample the same with your full-stained sample, using matched unstained cells) as a separate parameter largely improves unmixing. This applies to the cells that are not traditionally considered "autofluorescence heavy" too, for example, T cells. Dr Novo's team also has recent publication on their more complex unmixing methods. However, optimising experimental conditions upstream is the most important. OLS is the method used by software like *Spectroflo*.


