---
title: "Residual analysis"
format: html
editor: visual
---

```{r, include = FALSE}
library(tidyverse)
library(matrixStats)
library(corrplot)
library(data.table)
ggplot2::theme_set(theme_bw())
set.seed(233)
```


Author: Yuchen Li
Statistical consutation: Ryan Kmet, Dr David Novo, Elijah Willie.
Version: 2026 Apr

## Residual analysis

### plot the residuals

```{r, fig.width=15, fig.height=5}
# density plots
res <- t(res)
colnames(res)<-aurora_channels
res %>% 
  as.data.frame() %>% 
  pivot_longer(cols = colnames(res), names_to = "channel", values_to = "values") %>% 
ggplot(aes(x = values))+
  geom_histogram(stat = "density", col = "black", fill = "cornflowerblue")+
  scale_x_continuous(limits = c(-1e+05, 1e+05))+
  theme_bw()

res %>% 
  as.data.frame() %>% 
  pivot_longer(cols = colnames(res), names_to = "channel", values_to = "values") %>% 
ggplot(aes(x = values))+
  geom_density(linewidth = 1, fill = "cornflowerblue")+
  scale_x_continuous(limits = c(-1e+05, 1e+05))+
  theme_bw()

```

### r vs rhat

```{r, fig.width = 12, fig.height = 5}
# Mean
dim(r_fluo)
mean_r_fluo <- rowMeans(r_fluo)
mean_rhat <-rowMeans(r_hat)

names(mean_rhat) <- aurora_channels
names(mean_r_fluo) <- aurora_channels

r_rhat_mean <- cbind(mean_r_fluo, mean_rhat)
r_rhat_mean <- as.data.frame(r_rhat_mean)
r_rhat_mean$channel <- rownames(r_rhat_mean)
ggplot(r_rhat_mean, aes(x = mean_r_fluo, y = mean_rhat))+
  geom_point()+
    geom_text(aes(label = channel), hjust= 0.5,vjust = -0.5)+
  geom_smooth(aes(colour = NULL))+
  theme_bw()


# Median
med_r_fluo <- rowMedians(r_fluo)
med_rhat <- rowMedians(r_hat)

r_rhat_med <- cbind(med_r_fluo, med_rhat)

r_rhat_med <- as.data.frame(r_rhat_med)

r_rhat_med$channel <- aurora_channels

ggplot(r_rhat_med, aes(x = med_r_fluo, y = med_rhat))+
geom_point()+
  geom_text(aes(label = channel), hjust= 0.5,vjust = -0.5)+
  geom_smooth(aes(colour = NULL))+
  theme_bw()
r_rhat_med$residual <- med_r_fluo - med_rhat
sd(r_rhat_med$residual)
```

### RMSE and RMEdSE

```{r}
# RMSE of r and rhat
mse<- colMeans((res)^2)
rmse <- sqrt(mse)

# plot
# generate a df
df_rmse <- data.frame(channels = names(rmse), values = rmse)
df_rmse %>% 
  ggplot(aes(x = factor(channels, levels = aurora_channels),y = values))+
  geom_point()+
  theme_bw()+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.05, hjust = 0.3))

```

RMedSE

```{r}
medse <- colMedians((res)^2)
rmedse <- sqrt(medse)

# plot
df_rmedse <- data.frame(channels = names(rmedse), values = rmedse)
df_rmedse %>% 
  ggplot(aes(x = factor(channels, levels = aurora_channels),y = values))+
  geom_point()+
  theme_bw()+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.05, hjust = 0.3))
```

### scale-location plot

```{r}
# calculate standardized residuals
standard_res <- sqrt(abs(res/rowSds(res)))

# Generate the plot

scale_loc <- data.frame(
  channels = colnames(standard_res),
  standard_residuals = colMeans(standard_res),
  fitted_values = colMeans(r_hat)
)

ggplot(scale_loc, aes(x = fitted_values, y = standard_residuals))+
geom_point()+
  theme_bw()+
  scale_y_continuous(limits = c(0,1.8))

```

Residual plot

```{r, fig.width = 9, fig.height = 5}
# calculate median residuals
df_med_res <- data.frame(
  channels = colnames(res),
  values  = colMedians(res)
)

#residual plot
ggplot(df_med_res, aes(x = factor(channels, levels = aurora_channels), y = values))+
  geom_point()+
  theme_bw()+
  geom_hline(yintercept = 0)+
  theme(axis.text.x =element_text(angle = 90,face = "bold"))
ggsave("./plots/residual analysis/residual_plot_CUM500_ols.png",width = 18, height = 10, units = c("cm"))

```

Save the residual analysis results

```{r}
res_analysis_ols <- list(
  subsampled_r = r,
  r_fluoescence = r_fluo,
  r_expected = r_hat,
  scale_location = scale_loc,
  df_root_MedSE = df_rmedse,
  df_root_MSE = df_rmse,
  residuals = res,
  standardised_residuals = standard_res
)

saveRDS(res_analysis_ols, "./rds/20260202_residual_analysis_CUM500_ols.rds")

res_analysis_ols <- readRDS("./rds/20260202_residual_analysis_CUM500_ols.rds")
res <- res_analysis_ols$residuals

df_med_res <- data.frame(
  channels = colnames(res),
  values  = colMedians(res)
)

#residual plot
ggplot(df_med_res, aes(x = factor(channels, levels = aurora_channels), y = values))+
  geom_point(aes(size = 0.2))+
  theme_bw()+
  geom_hline(yintercept = 0)+
  theme(axis.text.x =element_text(angle = 90,face = "bold", vjust = 0.1))+
  theme(panel.border = element_rect(linewidth = 1))
ggsave("./plots/residual analysis/residual_plot_CUM500_ols.png",width = 18, height = 10, units = c("cm"))
```

### Residuals correlation

```{r, fig.width=15, fig.height=15}
res_analysis <- readRDS("./rds/20260202_residual_analysis_CUM500_ols.rds")
res <- res_analysis$residuals
cor_res <- cor(res)
corrplot(cor_res,method = "circle",tl.col = "black", type = c("low"))

```

## Calculate the variance, co-variance matrix

This method was described by (Kmet & Novo, 2025)\[<https://onlinelibrary.wiley.com/doi/10.1002/cyto.a.24957>\] and mentioned in our discussion with them. It is a quick diagnostic tool to identify where the issues in unmixing comes from large spread and variation in AF (elements on the diagonal) or potential dye interactions (off-diagonal elements).

### Calculate residuals

Model residuals are calculated by predicted total signals reconstructed by Ma - observed signals. M is the spectral profile matrix we calculated in the OLS unmixing section. a is the 'abundances' i.e. unmixed data. The below codes assumes that you have done OLS unmixing and saved the OLS unmixed results. If you don't have an object saved, please run through the OLS unmixing script first.

```{r}
# load the object
ols_unmix <- readRDS("./rds/OLS_unmixing.rds") 

# take the unmixed fluorescence channel data
a <- ols_unmix$Unmixed_fluo_channels

# Retrive the estimated total signal from Ma 
m <- ols_unmix$Spectral_matrix 
r_hat <- m%*%a  

# Calculate residuals 
res <- r_fluo - r_hat 
dim(res) #quick check
```


### Calculate residual variance and the (M^T M)^(-1) portion

I use *data.table*'s matrix transpose function cause it's faster on my laptop. 

```{r, fig.height=12, fig.width=18}
# calculate variance
var <- var(as.vector(re))

# calculate (M^T M)^(-1)
# Calculate M^T
m <- as.data.table(m)
m_t<- transpose(m)
m <- as.matrix(m)
m_t <- as.matrix(m_t)

# caclulate the variance, co-variance matrix
vcv <- var*mtm

# visualise
pheatmap(vcv, cluster_cols = F, cluster_rows = F, show_rownames = T)
corrplot(vcv)
```

## Investigate the inherent correlation between fluorophores using SS control profiles

```{r,fig.width =10, fig.height = 10}
spectral_matrix <- ols_unmix$Spectral_matrix
cor_m<- cor(spectral_matrix)
corrplot(cor_m, type = c("lower"), tl.col = "black")
```

## Investigate the residuals after unmixing with AF

In R, we can also calculate the classical model diagnostics than using software. For what it's worth: I calculated the residuals from unmixing without AF extraction, and found that the largest errors line perfectly with AF hotspots. The unmixing was largely improved by adding the unstained sample and the AF parameter, even though I'm not studying a typical "AF high" cell type.\ 

Below are a few examples for quick residual analysis.

### Quickly plot the residuals

This should be roughly approaching normal, as we have millions of events with potential Poisson noise (although not completely Poisson).

```{r, fig.width=15, fig.height=5}
# density plots with the Aurora channel order
res <- t(res)
aurora_channels <- readRDS("./rds/aurora_channels.rds")
colnames(res)<-aurora_channels
res %>% 
  as.data.frame() %>% 
  pivot_longer(cols = colnames(res), names_to = "channel", values_to = "values") %>% 
ggplot(aes(x = values))+
  geom_density(linewidth = 1, fill = "cornflowerblue")+
  scale_x_continuous(limits = c(-1e+05, 1e+05))+
  theme_bw()
```

### Scale-location plot

```{r}
# calculate standardized residuals
res <- as.data.table(res)
res <- transpose(res)
res<- as.matrix(res)
standard_res <- sqrt(abs(res/rowSds(res)))

# Generate the plot
scale_loc <- data.frame(
  channels = aurora_channels,
  standard_residuals = rowMeans(standard_res),
  fitted_values = rowMeans(r_hat)
)

ggplot(scale_loc, aes(x = fitted_values, y = standard_residuals))+
geom_point()+
  theme_bw()+
  scale_y_continuous(limits = c(0,1.8))
```

Residual plot.

```{r, fig.width = 9, fig.height = 5}
# calculate median residuals
df_med_res <- data.frame(
  channels = rownames(r_hat),
  values  = colMedians(res)
)

#residual plot
ggplot(df_med_res, aes(x = factor(channels, levels = aurora_channels), y = values))+
  geom_point()+
  theme_bw()+
  geom_hline(yintercept = 0)+
  theme(axis.text.x =element_text(angle = 90,face = "bold"))
ggsave("./plots/residual analysis/residual_plot_CUM500_ols_withAF.png",width = 18, height = 10, units = c("cm"))
```

### Investigate the residuals variances and channel MFI

```{r}
# Transpose residuals as the channels as columns
res <- as.data.table(res)
res <- transpose(res)
res<- as.matrix(res)
colnames(res) <- aurora_channels
res_variance <- colVars(res)
res_sd <- colSds(res)

# Load the raw sample fs data
fs <- ols_unmix$raw_sample_flowframe
expr <- exprs(fs)
expr <- expr[,aurora_channels]
expr_med <- colMedians(expr)
expr_mean <- colMeans(expr)
expr_sd <- colSds(expr)

#Calculate summary stats
df <- data.frame(Channel_mean = expr_mean,
                Channel_median = expr_med,
                Channel_SD = expr_sd,
                Channels = names(expr_mean),
                Residual_SD = res_sd,
                Residual_variance = res_variance)

# Plotting residuals against the channel signal intensities
ggplot(df, aes(x = Channel_mean, y = Residual_SD, colour = factor(Channels, levels = aurora_channels)))+
  geom_point(aes(size=0.5),show.legend = FALSE)+
  geom_smooth(method = "lm", colour = "blue")+
  theme_bw()+
  scale_color_viridis_d()+
  theme(panel.border = element_rect(linewidth = 1))

ggplot(df, aes(x = Channel_median, y = Residual_SD, colour = factor(Channels, levels = aurora_channels)))+
  geom_point(aes(size=0.5),show.legend = FALSE)+
  geom_smooth(method = "lm", colour = "red")+
  theme_bw()+
  scale_color_viridis_d()+
  theme(panel.border = element_rect(linewidth = 1))

cor(df$Residual_SD, df$Channel_mean)
cor(df$Residual_SD, df$Channel_median)

# Calculate sum of squared mean errors
mean_sse <- sum((colMeans(res))^2)
```

```{r}
# For a quick record
sessionInfo()
```

