---
title: "Residual analysis"
format: html
editor: visual
---

```{r, include = FALSE}
library(tidyverse)
library(matrixStats)
library(corrplot)
library(data.table)
library(flowCore)
ggplot2::theme_set(theme_bw())
set.seed(233)
```

Author: Yuchen Li Version: 2026 Apr

## Residual analysis

## Calculate the variance, co-variance matrix

This method was described by (Kmet & Novo, 2025)\[<https://onlinelibrary.wiley.com/doi/10.1002/cyto.a.24957>\] and mentioned in our discussion with them. It is a quick diagnostic tool to identify where the issues in unmixing comes from large spread and variation in AF (elements on the diagonal) or potential dye interactions (off-diagonal elements).

### Calculate residuals

Model residuals are calculated by predicted total signals reconstructed by Ma - observed signals. M is the spectral profile matrix. a is the 'abundances' i.e. unmixed data/coefficients from the linear system. The below codes assumes that you have done OLS unmixing and saved the OLS unmixed results. If you don't have an object saved, please run through the unmixing script first. Or calculate them from the scratch here.

```{r}
# load the object
unmix <- readRDS("./rds/xx_unmixing.rds") 

# take the unmixed fluorescence channel data
a <- unmix$Unmixed_fluo_channels

# Retrive the estimated total signal from Ma 
m <- unmix$Spectral_matrix 
r_hat <- m%*%a  

# Calculate residuals 
res <- r_fluo - r_hat 
dim(res) #quick check
```

### Calculate residuals from scratch if you did not have a master object saved

```{r}
# load data
#unmixed, a
unmixed <- readRDS("./rds/xx.rds")
a <- t(unmixed) # transpose it into the cells as columns vs channels as rows format
dim(a)
rownames(a)
colnames(M)
a <- a[rownames(a)%in%colnames(M),] # take only the fluorescence abundances from unmixed data

# M
M <- read.csv("spectral_matrix.csv")
M <- as.matrix(M)

# FS, r
fs_path <- "./full_stained_raw.fcs"
sample <- read.FCS(fs_path, transformation = "linearize-with-PnG-scaling")
#order of the cols
aurora_channels <- c('UV1-A',  'UV2-A',  'UV3-A',  'UV4-A',  'UV5-A',  'UV6-A',  'UV7-A',  'UV8-A',  'UV9-A',  'UV10-A','UV11-A', 'UV12-A', 'UV13-A', 'UV14-A', 'UV15-A', 'UV16-A', 'V1-A',   'V2-A',  
'V3-A',   'V4-A',   'V5-A',   'V6-A',   'V7-A',   'V8-A',   'V9-A',   'V10-A',  'V11-A',  'V12-A',  'V13-A',  'V14-A',  'V15-A',  'V16-A',  'B1-A',   'B2-A',   'B3-A',   'B4-A',  'B5-A',   'B6-A',   'B7-A',   'B8-A',   'B9-A',   'B10-A',  'B11-A',  'B12-A',  'B13-A',  'B14-A',  'YG1-A',  'YG2-A',  'YG3-A',  'YG4-A',  'YG5-A',  'YG6-A',  'YG7-A',  'YG8-A', 'YG9-A',  'YG10-A', 'R1-A',   'R2-A',   'R3-A',   'R4-A',   'R5-A',   'R6-A',   'R7-A',   'R8-A'  )
sample_fluo<-exprs(sample[,sample@parameters@data$name[sample@parameters@data$name %in% aurora_channels]])
dim(sample_fluo)
colnames(sample_fluo) <- NULL
# transpose it into the cells as columns vs channels as rows format
r <- t(sample_fluo)

# Retrive the estimated total signal from Ma
r_hat <- M%*%a
dim(r_hat)
# residuals
res <- r - r_hat
dim(res)
```

### Calculate residual variance and the (M\^T M)\^(-1) portion

I use *data.table*'s matrix transpose function when the matrix is large/strange, cause it's faster on my laptop.

```{r, fig.height=12, fig.width=18}
# calculate variance
var <- var(as.vector(res))

# calculate (M^T M)^(-1)
# Calculate M^T
m <- as.data.table(m)
m_t<- transpose(m)
m <- as.matrix(m)
m_t <- as.matrix(m_t)

mtm <- (M_t %*% M)^(-1)
# caclulate the variance, co-variance matrix
vcv <- var*mtm

# visualise
vcv[upper.tri(mtm)] <- NA

vcv %>% 
  reshape2::melt() %>% 
ggplot(aes(x = Var2, y = Var1, fill = value)) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_gradient2(
    low = "white",
    mid = "grey",
    high = "red",
    midpoint = 0,
    limits = c(0, max(mtm)),
    na.value = "transparent",
    oob = scales::squish    
  ) +
  coord_fixed() + # square tiles
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    panel.grid = element_blank()
  ) +
  labs(fill = "Value", x = NULL, y = NULL)
```

## Investigate the residuals after unmixing with AF

In R, we can also calculate the classical model diagnostics than using software. For what it's worth: I calculated the residuals from unmixing without AF extraction, and found that the largest errors line perfectly with AF hotspots. The unmixing was largely improved by adding the unstained sample and the AF parameter, even though I'm not studying a typical "AF high" cell type.

Below are a few examples for quick residual analysis.

### Quickly plot the residuals

This should be roughly approaching normal, as we have millions of events with potential Poisson noise (although not completely Poisson).

```{r, fig.width=15, fig.height=5}
# density plots with the Aurora channel order
res <- t(res) # columns as channels to plot
colnames(res)<-aurora_channels
res %>% 
  as.data.frame() %>% 
  pivot_longer(cols = colnames(res), names_to = "channel", values_to = "values") %>% 
ggplot(aes(x = values))+
  geom_density(linewidth = 1, fill = "cornflowerblue")+
  scale_x_continuous(limits = c(-1e+05, 1e+05))+
  theme_bw()
```

### Scale-location plot

```{r}
# transpose res back and calculate standardized residuals
res <- as.data.table(res)
res <- transpose(res)
res<- as.matrix(res)
standard_res <- sqrt(abs(res/rowSds(res)))

# Generate the plot
scale_loc <- data.frame(
  channels = aurora_channels,
  standard_residuals = rowMeans(standard_res),
  fitted_values = rowMeans(r_hat)
)

ggplot(scale_loc, aes(x = fitted_values, y = standard_residuals))+
geom_point()+
  theme_bw()+
  scale_y_continuous(limits = c(0,1.8))
```

Residual plot.

```{r, fig.width = 9, fig.height = 5}
# calculate median residuals
rownames(r) <- aurora_channels
rownames(res) <- aurora_channels
df_med_res <- data.frame(
  channels = rownames(r),
  values  = rowMedians(res)
)

#residual plot
ggplot(df_med_res, aes(x = factor(channels, levels = aurora_channels), y = values, size = 0.2))+
  scale_y_continuous(limits = c(-7000, 15000))+
  geom_point()+
  theme_bw()+
  geom_hline(yintercept = 0)+
  theme(axis.text.x =element_text(angle = 90,face = "bold"),
        panel.border = element_rect(linewidth = 1))
ggsave("./plots/residual analysis/xx.png",width = 18, height = 10, units = c("cm"))
```

### Investigate the residuals variances and channel MFI

```{r}
#Calculate summary stats
res_variance <- rowVars(res)
res_sd <- rowSds(res)
expr_med <- rowMedians(r)
expr_mean <- rowMeans(r)
expr_sd <- rowSds(r)

df <- data.frame(Channel_mean = expr_mean,
                Channel_median = expr_med,
                Channel_SD = expr_sd,
                Channels = names(expr_mean),
                Residual_SD = res_sd,
                Residual_variance = res_variance)

# Plotting residuals against the channel signal intensities
ggplot(df, aes(x = Channel_mean, y = Residual_SD, colour = factor(Channels, levels = aurora_channels)))+
  geom_point(aes(size=0.5),show.legend = FALSE)+
  geom_smooth(method = "lm", colour = "blue")+
  theme_bw()+
  scale_color_viridis_d()+
  theme(panel.border = element_rect(linewidth = 1))

ggplot(df, aes(x = Channel_median, y = Residual_SD, colour = factor(Channels, levels = aurora_channels)))+
  geom_point(aes(size=0.5),show.legend = FALSE)+
  geom_smooth(method = "lm", colour = "red")+
  theme_bw()+
  scale_color_viridis_d()+
  theme(panel.border = element_rect(linewidth = 1))

cor(df$Residual_SD, df$Channel_mean)
cor(df$Residual_SD, df$Channel_median)
```

### RMSE and RMEdSE

rmse across channels

```{r}
# RMSE of r and rhat
mse<- rowMeans((res)^2)
rmse <- sqrt(mse)

# plot
# generate a df
df_rmse <- data.frame(channels = names(rmse), values = rmse)
df_rmse %>% 
  ggplot(aes(x = factor(channels, levels = aurora_channels),y = values))+
  geom_point()+
  theme_bw()+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.05, hjust = 0.3))

```

RMedSE across channels

```{r}
medse <- rowMedians((res)^2)
rmedse <- sqrt(medse)

# plot
df_rmedse <- data.frame(channels = names(rmedse), values = rmedse)
df_rmedse %>% 
  ggplot(aes(x = factor(channels, levels = aurora_channels),y = values))+
  geom_point()+
  theme_bw()+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.05, hjust = 0.3))
```

```{r}
# For a quick record
sessionInfo()
```
