---
title: "Residual analysis"
format: html
editor: visual
---

```{r, include = FALSE}
library(tidyverse)
library(matrixStats)
library(data.table)
library(flowCore)
library(flowSpectrum)
ggplot2::theme_set(theme_bw())
set.seed(233)
```

Author: Yuchen Li Version: 2026 August

## Residual analysis

### Calculate residuals

Model residuals are calculated by observed raw signals-Ma. M is the spectral profile matrix. a is the 'abundances' i.e. unmixed data/coefficients from the linear system. The below codes assumes that you have done OLS unmixing and saved the OLS unmixed results. If you don't have an object saved, please run through the unmixing script first.

```{r}
# load the object
unmix <- readRDS("./rds/ols_unmixed.rds") 

# take the unmixed fluorescence channel data, transpose it to fluorophores as rows
# a <- unmix[, whichever columns]
a <- as.data.table(a)
a <- transpose(a)
a <- as.matrix(a)

# Retrive the estimated total signal from Ma 

# retrieve M from the ref controls
file.dir <-"./ref controls"
files <- dir(file.dir, pattern = "fcs$", full.names=T)
as.matrix(files)

# read in fcs files
dat <- lapply(files, read.FCS)
names(dat) <-gsub("./ref controls/","",files)
as.matrix(names(dat))
p1 <- lapply(seq(2,length(dat), by = 2), function(i){spectralplot(dat[[i]], normalize = TRUE, guessPop = TRUE, unstained = dat[[i-1]])$data})

#Name the list of new dfs
 names(p1) <- gsub("./ref control/", "",names(dat[seq(2,length(dat),2)]))
 names(p1) <- gsub("_pos.fcs", "",names(dat[seq(2,length(dat),2)]))
 print(as.matrix(names(p1)))
 
aurora_channels <- c('UV1-A',  'UV2-A',  'UV3-A',  'UV4-A',  'UV5-A',  'UV6-A',  'UV7-A',  'UV8-A',  'UV9-A',  'UV10-A','UV11-A', 'UV12-A', 'UV13-A', 'UV14-A', 'UV15-A', 'UV16-A', 'V1-A',   'V2-A',  
'V3-A',   'V4-A',   'V5-A',   'V6-A',   'V7-A',   'V8-A',   'V9-A',   'V10-A',  'V11-A',  'V12-A',  'V13-A',  'V14-A',  'V15-A',  'V16-A',  'B1-A',   'B2-A',   'B3-A',   'B4-A',  'B5-A',   'B6-A',   'B7-A',   'B8-A',   'B9-A',   'B10-A',  'B11-A',  'B12-A',  'B13-A',  'B14-A',  'YG1-A',  'YG2-A',  'YG3-A',  'YG4-A',  'YG5-A',  'YG6-A',  'YG7-A',  'YG8-A', 'YG9-A',  'YG10-A', 'R1-A',   'R2-A',   'R3-A',   'R4-A',   'R5-A',   'R6-A',   'R7-A',   'R8-A'  )

# Reorder the channels
for (i in 1:length(p1)){
  p1[[i]] <- p1[[i]][match(aurora_channels, p1[[i]]$`rownames(df)`),]
}
 
# add a sample column for plotting
  for(i in 1:length(p1)){p1[[i]]$sample = names(p1)[i]}
 
 cvars<-gsub("\ ","",apply(expand.grid(c("UV", 'V', 'B', "YG", "R"),1:16,"-A") %>% 
                            mutate(Var1=factor(Var1,levels=unique(Var1))) %>% 
                            arrange(Var1,Var2),1,function(X) paste0(X,collapse="")))

# combine the df into a big df
df <- df <- dplyr::bind_rows(c(p1))

## Import AF profile
dir <- "./af"
files <- dir(dir, pattern = "fcs$", full.names=T)
as.matrix(files)

# read in fcs files
dat <- lapply(files, read.FCS)

# Extract normalised spectra
af <- lapply(1:length(dat), function(i){spectralplot(dat[[i]], normalize = TRUE)$data})
names(af) <- names(dat)

# Reorder the channels
af <- af[[1]]
af<- af[match(aurora_channels, af$`rownames(df)`),]
af$sample <- "AF"

# attaching AF profile to the spectral profile data frame and list
df <- rbind(df,af)
p1$AF <- af

# construct the spectral signature matrix
Mlist <- list()

for (i in 1:length(table(df$sample))){
    Mlist[[i]] <- as.matrix(p1[[i]]$medians.max.medians.)
}

spectra_matrix <- do.call(cbind, Mlist)
rm(Mlist)
colnames(spectra_matrix) <- names(p1)

# Construct theorectical total signals
m <- spectra_matrix 
r_hat <- m%*%a  

# retrieve the raw observed signals, r_fluo
fs_path <- "./FS.fcs"
sample <- read.FCS(fs_path, transformation = "linearize-with-PnG-scaling")
sample_fluo<-exprs(sample[,sample@parameters@data$name[sample@parameters@data$name %in% df$`rownames(df)`]])
dim(sample_fluo)
colnames(sample_fluo) <- NULL

# transpose it into the cells as columns vs channels as rows format
sample_fluo<- as.data.table(sample_fluo)
sample_fluo<- transpose(sample_fluo)
sample_fluo <- as.matrix(sample_fluo)

r_fluo <- sample_fluo

# Calculate residuals 
res <- r_fluo - r_hat 
dim(res) #quick check
```

In R, we can also calculate the classical model diagnostics than using software. For what it's worth: I calculated the residuals from unmixing without AF extraction, and found that the largest errors line perfectly with AF hotspots. The unmixing was largely improved by adding the unstained sample and the AF parameter, even though I'm not studying a typical "AF high" cell type.

Below are a few examples for quick residual analysis.

### Quickly plot the residuals

This should be roughly approaching normal, as we have millions of events with potential Poisson noise (although not completely Poisson).

```{r, fig.width=15, fig.height=5}
# density plots with the Aurora channel order
res <- t(res) # columns as channels to plot
colnames(res)<-aurora_channels
res %>% 
  as.data.frame() %>% 
  pivot_longer(cols = colnames(res), names_to = "channel", values_to = "values") %>% 
ggplot(aes(x = values))+
  geom_density(linewidth = 1, fill = "cornflowerblue")+
  scale_x_continuous(limits = c(-1e+05, 1e+05))+
  theme_bw()
```

### Scale-location plot

```{r}
# transpose res back and calculate standardized residuals
res <- as.data.table(res)
res <- transpose(res)
res<- as.matrix(res)
standard_res <- sqrt(abs(res/rowSds(res)))

# Generate the plot
scale_loc <- data.frame(
  channels = aurora_channels,
  standard_residuals = rowMeans(standard_res),
  fitted_values = rowMeans(r_hat)
)

ggplot(scale_loc, aes(x = fitted_values, y = standard_residuals))+
geom_point()+
  theme_bw()+
  scale_y_continuous(limits = c(0,1.8))
```

### Median channel-wise residual plot.

```{r, fig.width = 9, fig.height = 5}
# calculate median residuals
colnames(res) <- aurora_channels
df_med_res <- data.frame(
  channels = colnames(res),
  values  = colMedians(res)
)

#residual plot
ggplot(df_med_res, aes(x = factor(channels, levels = aurora_channels), y = values))+
  geom_point()+
  theme_bw()+
  geom_hline(yintercept = 0)+
  theme(axis.text.x =element_text(angle = 90,face = "bold"),
        panel.border = element_rect(linewidth = 1))
#ggsave("./plots/residual analysis/xx.png",width = 18, height = 10, units = c("cm"))
```

Scaled median residual signature.

```{r}
#0-1 scaling 
res_scale <- (df_med_res$values-min(df_med_res$values))/(max(df_med_res$values)-min(df_med_res$values))  
res_scale_df <- data.frame(   values = res_scale,   detectors = 1:64 )  
#scaled plot 
ggplot(res_scale_df, aes(x = detectors, y = values))+  
  geom_line(linewidth = 0.5)+   
  theme_bw()+   
  geom_hline(yintercept = 0)+   
  theme(axis.text.x =element_text(angle = 0,face = "bold"),         
        panel.border = element_rect(linewidth = 1))+   
  xlab("Detectors")+   
  ylab("Scaled residual signature")
```

### Mean channel-wise residual plot

```{r}
# mean residuals
colnames(res) <- aurora_channels
df_mu_res <- data.frame(
  channels = colnames(res),
  values  = colMeans(res)
)

#residual plot
ggplot(df_mu_res, aes(x = factor(channels, levels = aurora_channels), y = values))+
  geom_point()+
  theme_bw()+
  geom_hline(yintercept = 0)+
  theme(axis.text.x =element_text(angle = 90,face = "bold"),
        panel.border = element_rect(linewidth = 1))

# scaled signature
mu_res_scale <- (df_mu_res$values-min(df_mu_res$values))/(max(df_mu_res$values)-min(df_mu_res$values))  
mu_res_scale_df <- data.frame(   values = mu_res_scale,   detectors = 1:64 )  
#scaled plot 
ggplot(mu_res_scale_df, aes(x = detectors, y = values))+  
  geom_line(linewidth = 0.5)+   
  theme_bw()+   
  geom_hline(yintercept = 0)+   
  theme(axis.text.x =element_text(angle = 0,face = "bold"),         
        panel.border = element_rect(linewidth = 1))+   
  xlab("Detectors")+   
  ylab("Scaled residual signature")

```

Calculate expected noise floor at the intensity in that channel and plotting

```{r}
mu_detector <- rowMeans(r_fluo) # assuming that var(signal) = k*mu(detector) as Poisson noise, when k=1
floor <- sqrt(mu_detector) # as sd^2 ~= k*mu(detector)
df_floor <- data.frame(
  channels = colnames(res),
  values  = floor
)

ggplot(df_floor, aes(x = factor(channels, levels = aurora_channels), y = values))+
  geom_point()+
  theme_bw()+
  geom_hline(yintercept = 0)+
  theme(axis.text.x =element_text(angle = 90,face = "bold"),
        panel.border = element_rect(linewidth = 1))

```

Scaled signature.

```{r}
#0-1 scaling 
sd_scale <- (df_floor$values-min(df_floor$values))/(max(df_floor$values)-min(df_floor$values))  

sd_scale_df <- data.frame(values = res_scale,detectors = 1:64)  
#scaled plot 
ggplot(sd_scale_df, aes(x = detectors, y = values))+   
  geom_line(linewidth = 0.5)+   
  theme_bw()+   
  geom_hline(yintercept = 0)+   
  theme(axis.text.x =element_text(angle = 0,face = "bold"),         
        panel.border = element_rect(linewidth = 1))+   
  xlab("Detectors")+   
  ylab("Scaled SNR signature")
```

## What's next

We are aware that the noise from the detectors should be Poisson - i.e. higher variance when higher signal intensity. But how much of the residuals propagate from the fluorophore signature matrix, how much arises when unmixing the full-stained sample? Which component contributes to noise? And how to evaluate the performance of the unmixing models mathematically so we can improve it? That's what I'm working on atm.

```{r}
# For a quick record
sessionInfo()
```
