---
title: "Residual analysis"
format: html
editor: visual
---

```{r, include = FALSE}
library(tidyverse)
library(matrixStats)
library(data.table)
library(flowCore)
ggplot2::theme_set(theme_bw())
set.seed(233)
```

Author: Yuchen Li Version: 2026 Apr

## Residual analysis

### Calculate residuals

Model residuals are calculated by predicted total signals reconstructed by Ma - observed signals. M is the spectral profile matrix. a is the 'abundances' i.e. unmixed data/coefficients from the linear system. The below codes assumes that you have done OLS unmixing and saved the OLS unmixed results. If you don't have an object saved, please run through the unmixing script first.

```{r}
# load the object
unmix <- readRDS("./rds/xx_unmixing.rds") 

# take the unmixed fluorescence channel data
a <- unmix$Unmixed_fluo_channels

# Retrive the estimated total signal from Ma 
m <- unmix$Spectral_matrix 
r_hat <- m%*%a  

# Calculate residuals 
res <- r_fluo - r_hat 
dim(res) #quick check
```

In R, we can also calculate the classical model diagnostics than using software. For what it's worth: I calculated the residuals from unmixing without AF extraction, and found that the largest errors line perfectly with AF hotspots. The unmixing was largely improved by adding the unstained sample and the AF parameter, even though I'm not studying a typical "AF high" cell type.

Below are a few examples for quick residual analysis.

### Quickly plot the residuals

This should be roughly approaching normal, as we have millions of events with potential Poisson noise (although not completely Poisson).

```{r, fig.width=15, fig.height=5}
# density plots with the Aurora channel order
res <- t(res) # columns as channels to plot
colnames(res)<-aurora_channels
res %>% 
  as.data.frame() %>% 
  pivot_longer(cols = colnames(res), names_to = "channel", values_to = "values") %>% 
ggplot(aes(x = values))+
  geom_density(linewidth = 1, fill = "cornflowerblue")+
  scale_x_continuous(limits = c(-1e+05, 1e+05))+
  theme_bw()
```

### Scale-location plot

```{r}
# transpose res back and calculate standardized residuals
res <- as.data.table(res)
res <- transpose(res)
res<- as.matrix(res)
standard_res <- sqrt(abs(res/rowSds(res)))

# Generate the plot
scale_loc <- data.frame(
  channels = aurora_channels,
  standard_residuals = rowMeans(standard_res),
  fitted_values = rowMeans(r_hat)
)

ggplot(scale_loc, aes(x = fitted_values, y = standard_residuals))+
geom_point()+
  theme_bw()+
  scale_y_continuous(limits = c(0,1.8))
```

Median channel-wise residual plot.

```{r, fig.width = 9, fig.height = 5}
# calculate median residuals
rownames(r) <- aurora_channels
rownames(res) <- aurora_channels
df_med_res <- data.frame(
  channels = rownames(r),
  values  = rowMedians(res)
)

#residual plot
ggplot(df_med_res, aes(x = factor(channels, levels = aurora_channels), y = values, size = 0.2))+
  scale_y_continuous(limits = c(-7000, 15000))+
  geom_point()+
  theme_bw()+
  geom_hline(yintercept = 0)+
  theme(axis.text.x =element_text(angle = 90,face = "bold"),
        panel.border = element_rect(linewidth = 1))
ggsave("./plots/residual analysis/xx.png",width = 18, height = 10, units = c("cm"))
```

## What's next

We are aware that the noise from the detectors should be Poisson - i.e. higher variance when higher signal intensity. But how much of the residuals propagate from the fluorophore signature matrix, how much arises when unmixing the full-stained sample? And how to evaluate the performance of the unmixing models mathematically so we can improve it? That's what I'm working on atm.

```{r}
# For a quick record
sessionInfo()
```
