```{r setup, include=FALSE}
library(flowCore)
library(tidyverse)
library(flowSpectrum)
library(ggcyto)
library(pheatmap)
library(matrixStats)
library(corrplot)
library(data.table)
set.seed(233)
library(reticulate)
```

```{python}
import numpy as np
from sklearn import linear_model
from sklearn.model_selection import train_test_split, GridSearchCV
from sklearn.metrics import mean_squared_error
import pandas as pd
import fcsparser
np.random.seed(233)
```

Ridge regression handles the noisy data better (has co-linearity, and noisy), and it adds L2 panelties to marker abundances to shrink the extremely large signals. I found this method correcting the worst spillover issue in my own panel and removing extreme noise. So I would encourage you to give it a go. Make sure that you include the AF parameter so that the model is complete.

The main parameter that controls the strength of this shrinkage is the alpha. Here, I started with some random small alphas and let the machine pick the best one using a standard folded cross-validation. Feel free to experiment different start-up alphas and inspect your data after the tweaks.

Before start, build the spectral matrix then convert your spectral matrix from R into a .csv to load in as a python object.

```{r}
# load reference controls
file.dir <-"./ref controls"
files <- dir("./ref controls", pattern = "fcs$", full.names=T)

# read in fcs files
dat <- lapply(files, read.FCS)
names(dat) <-gsub("./ref controls/","",files)

p1 <- lapply(seq(2,length(dat), by = 2), function(i){spectralplot(dat[[i]], normalize = TRUE, guessPop = TRUE, unstained = dat[[i-1]])$data})

#Name the list of new dfs
 names(p1) <- gsub("./ref control/", "",names(dat[seq(2,length(dat),2)]))
 names(p1) <- gsub("_pos.fcs", "",names(dat[seq(2,length(dat),2)]))
 print(as.matrix(names(p1)))

### If using flowjo exported files: switch to this exact order of channels.

aurora_channels <- c('UV1-A',  'UV2-A',  'UV3-A',  'UV4-A',  'UV5-A',  'UV6-A',  'UV7-A',  'UV8-A',  'UV9-A',  'UV10-A','UV11-A', 'UV12-A', 'UV13-A', 'UV14-A', 'UV15-A', 'UV16-A', 'V1-A',   'V2-A',  'V3-A',   'V4-A',   'V5-A',   'V6-A',   'V7-A',   'V8-A',   'V9-A',   'V10-A',  'V11-A',  'V12-A',  'V13-A',  'V14-A',  'V15-A',  'V16-A',  'B1-A',   'B2-A',   'B3-A',   'B4-A',  'B5-A',   'B6-A',   'B7-A',   'B8-A',   'B9-A',   'B10-A',  'B11-A',  'B12-A',  'B13-A',  'B14-A',  'YG1-A',  'YG2-A',  'YG3-A',  'YG4-A',  'YG5-A',  'YG6-A',  'YG7-A',  'YG8-A', 'YG9-A',  'YG10-A', 'R1-A',   'R2-A',   'R3-A',   'R4-A',   'R5-A',   'R6-A',   'R7-A',   'R8-A'  )

# You can also save the aurora_channels.rds for future easy use.

# Reorder the channels
for (i in 1:length(p1)){
  p1[[i]] <- p1[[i]][match(aurora_channels, p1[[i]]$`rownames(df)`),]
}

#### fetch the channel names directly if you export data from cytek software
 cvars<-gsub("\ ","",apply(expand.grid(c("UV", 'V', 'B', "YG", "R"),1:16,"-A") %>% 
                            mutate(Var1=factor(Var1,levels=unique(Var1))) %>% 
                            arrange(Var1,Var2),1,function(X) paste0(X,collapse="")))

# add a sample column for plotting
  for(i in 1:length(p1)){p1[[i]]$sample = names(p1)[i]}

# combine the df into a big df
df <- df <- dplyr::bind_rows(c(p1))

# Add the AF parameter in
dir <- "./af"
files <- dir(dir, pattern = "fcs$", full.names=T)
as.matrix(files)

# read in fcs files
dat <- lapply(files, read.FCS)

# tidy up the file names as you wish, if you need to
#names(dat) <-gsub("./af/export_","",files)
#names(dat) <-gsub("_2026.*","",names(dat))

# Extract normalised spectra
af <- lapply(1:length(dat), function(i){spectralplot(dat[[i]], normalize = TRUE)$data})
names(af) <- names(dat)

# Reorder the channels
af<- af[match(aurora_channels, af$`rownames(df)`),]

### construct the unmixing matrix with AF
# add AF to the control set
spectra_matrix <-as.data.frame(spectra_matrix)
spectra_matrix$AF <- af$`medians.max.medians.`

write.csv(spactra_matrix,"spectral_matrix.csv")
```

### Load data

```{python}
# read in the raw full-stained data
path = "path/to/full-stained raw data.fcs"

# parse both metadata and data, get the expression values into a data frame
meta, data = fcsparser.parse(path, meta_data_only=False, reformat_meta=True)

# have a look
print(type(data))
data

# read in the spectral matrix
M = pd.read_csv("./spectral_matrix.csv")
print(type(M))

# take only the fluorescence channels from the raw data
Y = data.filter(regex="(UV|V|B|YG|R)(\d+)")

# transpose to have channels as rows
YT = Y.T

# store the non fluorecence channel of raw data
C = data.drop(columns=Y.columns)

# transpose to have channels as row names
YT = Y.T

# final checks
print(YT.shape)
print(M.shape)
print(type(M))
print(type(YT))
print(M.columns)
```

### Build the model

```{python}
# train-test split
x_train, x_test, y_train, y_test = train_test_split(M, YT, test_size=0.25,
                                                    random_state=42)

# pick alpha using CV and build the best model based on smallest MSE
param_grid={"alpha":[1e-3,1e-2,0.1, 0.2, 0.5]} # make a param_grid dic
ridge_grid = GridSearchCV(
    linear_model.Ridge(solver='svd'),
    param_grid,
    cv=5,
    scoring
    ="neg_mean_squared_error"
)
ridge_grid.fit(x_train, y_train)
best_ridge=ridge_grid.best_estimator_ # store the best pipeline based on the alpha picked by CV

# take the best model's coefficients
best_A = best_ridge.coef_

# Look at what the machine reckons the best alpha, and if the data is in a right dimension
print(best_A.shape)
print(best_ridge.alpha)

# generate a df
feature_names = M.columns

best_A = pd.DataFrame(
    best_ridge.coef_,
    columns=feature_names
)

# Stick it back to the untouched non fluorescence channel data
unmix = pd.concat([best_A, C], axis=1)

# Write to .csv
unmix.to_csv("ridge_unmixed_sample_name.csv", index=False)

```

```{r}
# process and write fcs
a <- read.csv("ridge_unmixed_sample_name.csv")
a<- as.matrix(a)
a_ff<-flowFrame(a)
write.FCS(a_ff, "ridge_unmixed_sample_name.fcs")
# Save your results
saveRDS(a, "ridge_unmixed_sample_name.rds")
gc()
```

From here, you can either proceed with my residual_analysis.qmd to analyse the model performance, or you can have a quick look at the model errors using the test set. From my experience, you would always get a large number here as the total MSE. This is just the nature of flow data (values vary across a few logs). So as long as you think the unmixing is okay during manual inspection, you are still good.

```{python}
# Look at the errors
y_pred = best_ridge.predict(x_test)
print(mean_squared_error(y_test, y_pred))
```

Note: I put python and R codes interchangbly for an all-in-one workflow. But if you have VS Code and are comfortable at using it, I still suggest you to use that for running python codes then jump back to Rstudio to run all those data format conversions. Python kernel has to go through reticulate in quarto documents, so it's slower to run here. VS Code also has autofilling for python codes if you like it.
