```{r setup, include=FALSE}
library(flowCore)
library(tidyverse)
library(flowSpectrum)
library(ggcyto)
library(pheatmap)
library(matrixStats)
library(corrplot)
library(data.table)
set.seed(233)
library(reticulate)
```


Ridge regression handles the noisy data better (has co-linearity, and ill-conditioned), and it adds L2 panelties to marker abundances to shrink the extremely large signals. I found this method correcting the worst spillover issue in my own panel and removing extreme noise. So I would encourage you to give it a go. Make sure that you include the AF parameter so that the model is complete.

The main parameter that controls the strength of this shrinkage is the alpha. Here, I started with some random small alphas and let the machine pick the best one using a standard folded cross-validation. Feel free to experiment different start-up alphas and inspect your data after the tweaks.

Before start, build the spectral matrix then convert your spectral matrix from R into a .csv to load in as a python object.

```{r}
# load controls
file.dir <-"./ref controls"
files <- dir(file.dir, pattern = "fcs$", full.names=T)
as.matrix(files)

# read in fcs files
dat <- lapply(files, read.FCS)
names(dat) <-gsub("./ref controls/","",files)
as.matrix(names(dat))


p1 <- lapply(seq(2,length(dat), by = 2), function(i){spectralplot(dat[[i]], normalize = TRUE, guessPop = TRUE, unstained = dat[[i-1]])$data})

#Name the list of new dfs
 names(p1) <- gsub("./ref control/", "",names(dat[seq(2,length(dat),2)]))
 names(p1) <- gsub("_pos.fcs", "",names(dat[seq(2,length(dat),2)]))
 print(as.matrix(names(p1)))
 
aurora_channels <- c('UV1-A',  'UV2-A',  'UV3-A',  'UV4-A',  'UV5-A',  'UV6-A',  'UV7-A',  'UV8-A',  'UV9-A',  'UV10-A','UV11-A', 'UV12-A', 'UV13-A', 'UV14-A', 'UV15-A', 'UV16-A', 'V1-A',   'V2-A',  
'V3-A',   'V4-A',   'V5-A',   'V6-A',   'V7-A',   'V8-A',   'V9-A',   'V10-A',  'V11-A',  'V12-A',  'V13-A',  'V14-A',  'V15-A',  'V16-A',  'B1-A',   'B2-A',   'B3-A',   'B4-A',  'B5-A',   'B6-A',   'B7-A',   'B8-A',   'B9-A',   'B10-A',  'B11-A',  'B12-A',  'B13-A',  'B14-A',  'YG1-A',  'YG2-A',  'YG3-A',  'YG4-A',  'YG5-A',  'YG6-A',  'YG7-A',  'YG8-A', 'YG9-A',  'YG10-A', 'R1-A',   'R2-A',   'R3-A',   'R4-A',   'R5-A',   'R6-A',   'R7-A',   'R8-A'  )

# Reorder the channels
for (i in 1:length(p1)){
  p1[[i]] <- p1[[i]][match(aurora_channels, p1[[i]]$`rownames(df)`),]
}
 
# add a sample column for plotting
  for(i in 1:length(p1)){p1[[i]]$sample = names(p1)[i]}
 
 cvars<-gsub("\ ","",apply(expand.grid(c("UV", 'V', 'B', "YG", "R"),1:16,"-A") %>% 
                            mutate(Var1=factor(Var1,levels=unique(Var1))) %>% 
                            arrange(Var1,Var2),1,function(X) paste0(X,collapse="")))

# combine the df into a big df
df <- df <- dplyr::bind_rows(c(p1))

## Import AF profile
dir <- "./af"
files <- dir(dir, pattern = "fcs$", full.names=T)
as.matrix(files)

# read in fcs files
dat <- lapply(files, read.FCS)

# tidy up the file names as you wish, if you need to
#names(dat) <-gsub("./af/export_","",files)
#names(dat) <-gsub("_2026.*","",names(dat))

# Extract normalised spectra
af <- lapply(1:length(dat), function(i){spectralplot(dat[[i]], normalize = TRUE)$data})
names(af) <- names(dat)

# Reorder the channels
af <- af[[1]]
af<- af[match(aurora_channels, af$`rownames(df)`),]
af$sample <- "AF"

# attaching AF profile to the spectral profile data frame and list
df <- rbind(df,af)
p1$AF <- af


### construct the spectra matrix
Mlist <- list()

for (i in 1:length(table(df$sample))){
    Mlist[[i]] <- as.matrix(p1[[i]]$medians.max.medians.)
}
spectra_matrix <- do.call(cbind, Mlist)
rm(Mlist)

## write it out for python to use
spectra_matrix <-as.data.frame(spectra_matrix)
colnames(spectra_matrix)<- names(p1)
write.csv(spectra_matrix,"spectral_matrix.csv", row.names = FALSE)
```

### SVD and calculate ridge candidates if you are not sure which one to use

```{r}
# SVD of the spectra matrix
M <- as.matrix(spectra_matrix)
svd_M <- svd(M)
d <- svd_M$d
# Make ridge candidates from the data
min(d)
log10(min(d))
max(d)
log10(max(d))
lg_d <- log10(d)
lg_d

lg_d[0 < lg_d & lg_d <0.5] # 0 is the baseline of OLS. We don't wanna -ve regularisation, don't want the large ones that push the data points towards 0 too much either.
alpha_cand <- min(lg_d[0 < lg_d & lg_d <0.1])
# from here, you can also add some even smaller ones like 0.01, 0.005, 0.001 for the machine to try
```

```{python}
import numpy as np
from sklearn import linear_model
from sklearn.linear_model import RidgeCV
from sklearn.metrics import mean_squared_error
import pandas as pd
import fcsparser
np.random.seed(233)
```

### Load data

```{python}
# read in the raw full-stained data
path = "path_to_fullstain_raw.fcs"

# parse both metadata and data, get the expression values into a data frame
meta, data = fcsparser.parse(path, meta_data_only=False, reformat_meta=True)

# have a look
print(type(data))
data

# take only the fluorescence channels from the raw data
Y = data.filter(regex="(UV|V|B|YG|R)(\d+)")

# transpose to have channels as rows
YT = Y.T

# store the non fluorecence channel of raw data
C = data.drop(columns=Y.columns)

# transpose to have channels as row names
YT = Y.T

# read in the spectral matrix
M = pd.read_csv("./spectral_matrix.csv") # If you saved the spectral matrix as .csv from the last step.
#M = r['spectra_matrix'] # Access r objects in qmd py env
print(type(M))

# ridge alpha candidates
alpha_cand = r[0,1,0.01,0.001,0.0001]

```

```{python}
# final checks
print(YT.shape)
print(M.shape)
print(type(M))
print(type(YT))
print(M.columns)
```


### Build the model

```{python}

# Randomly selecte 75% of the channels to train the model to avoid overfitting. It performs better than using all data to train it. The model will then be applied to everything.
x_train, x_test, y_train, y_test = train_test_split(M, YT, test_size=0.25, random_state=42)

# pick alpha using CV and build the best model based on smallest MSE
#Candidates generated by smallest positive ones from the lg of the d from SVD of my spectra matrix. Then added some small values around it on a log scale.
m1 = linear_model.RidgeCV(alphas=alpha_cand) # use LOO CV by default, with negative MSE as loss function by default, if you have strictly positive values or are comfortable with directly shift your raw data to positives, you can use poisson deviance.

m1.fit(x_train, y_train)
# take the best model's coefficients
print(m1.alpha_)
best_a = m1.coef_
print(best_a.shape)

# generate a df
feature_names = M.columns
best_A = pd.DataFrame(
    m1.coef_,
    columns=feature_names
)
# stitch it back to the untouched non-fluorescence channels
unmix = pd.concat([best_A, C], axis=1)

# Write to .csv
unmix.to_csv("name_unmixed.csv", index=False)
```

Exit python and proceed from R. 

```{r}
# process and write fcs
a <- read.csv("./name_unmixed.csv")
a<- as.matrix(a)
a_ff<-flowFrame(a)
write.FCS(a_ff, "name_unmixed.fcs")
# Save your results
saveRDS(a, "./rds/name_unmixed.rds")
gc()
```

From here, you can either proceed with my residual_analysis.qmd to analyse the model performance, or you can have a quick look at the model errors using the test set. From my experience, you would always get a large number here as the total MSE. This is just the nature of flow data (values vary across a few logs). So as long as you think the unmixing is okay during manual inspection, you are still good.

```{python}
# Look at the errors
y_pred = best_ridge.predict(x_test)
print(mean_squared_error(y_test, y_pred))
```

Note: I put python and R codes interchangbly for an all-in-one workflow. But if you have VS Code and are comfortable at using it, I still suggest you to use that for running python codes then jump back to Rstudio to run all those data format conversions. Python kernel has to go through reticulate in quarto documents, so it's slower to run here. It also caused crashes in R sessions on my small laptop.


