---
title: "Unmixing in R"
author: "Yuchen Li"
date: "2026-0818"
output: 
  html_document:
    code_folding: "hide"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(flowCore)
library(flowSpectrum)
library(ggcyto)
library(pheatmap)
library(matrixStats)
library(data.table)
set.seed(233)
# dir.create("./rds")
# dir.create("./plots")
```

## Load controls

## Yuchen: Bypassing the pre-set automatic gating functions in AutoSpectral and use pregated fcs files from SpectroFlo. To generate the ref control spectral profiles

### Load pre-gated fcs files for reference controls 

```{r, message=FALSE, warning=FALSE}
file.dir <-"./ref controls"
files <- dir("./ref controls", pattern = "fcs$", full.names=T)
as.matrix(files)

# read in fcs files
dat <- lapply(files, read.FCS)
names(dat) <-gsub("./ref controls/","",files)
as.matrix(names(dat))
```

## Generate the data frame of reference control spectral profiles

```{r, message=FALSE, warning=FALSE, echo=FALSE}
p1 <- lapply(seq(2,length(dat), by = 2), function(i){spectralplot(dat[[i]], normalize = TRUE, guessPop = TRUE, unstained = dat[[i-1]])$data})

#Name the list of new dfs
 names(p1) <- gsub("./ref control/", "",names(dat[seq(2,length(dat),2)]))
 names(p1) <- gsub("_pos.fcs", "",names(dat[seq(2,length(dat),2)]))
 print(as.matrix(names(p1)))
 
 
 aurora_channels <- c('UV1-A',  'UV2-A',  'UV3-A',  'UV4-A',  'UV5-A',  'UV6-A',  'UV7-A',  'UV8-A',  'UV9-A',  'UV10-A','UV11-A', 'UV12-A', 'UV13-A', 'UV14-A', 'UV15-A', 'UV16-A', 'V1-A',   'V2-A',  
'V3-A',   'V4-A',   'V5-A',   'V6-A',   'V7-A',   'V8-A',   'V9-A',   'V10-A',  'V11-A',  'V12-A',  'V13-A',  'V14-A',  'V15-A',  'V16-A',  'B1-A',   'B2-A',   'B3-A',   'B4-A',  'B5-A',   'B6-A',   'B7-A',   'B8-A',   'B9-A',   'B10-A',  'B11-A',  'B12-A',  'B13-A',  'B14-A',  'YG1-A',  'YG2-A',  'YG3-A',  'YG4-A',  'YG5-A',  'YG6-A',  'YG7-A',  'YG8-A', 'YG9-A',  'YG10-A', 'R1-A',   'R2-A',   'R3-A',   'R4-A',   'R5-A',   'R6-A',   'R7-A',   'R8-A'  )

# Reorder the channels
for (i in 1:length(p1)){
  p1[[i]] <- p1[[i]][match(aurora_channels, p1[[i]]$`rownames(df)`),]
}
 
 
 
# add a sample column for plotting
  for(i in 1:length(p1)){p1[[i]]$sample = names(p1)[i]}
 
 cvars<-gsub("\ ","",apply(expand.grid(c("UV", 'V', 'B', "YG", "R"),1:16,"-A") %>% 
                            mutate(Var1=factor(Var1,levels=unique(Var1))) %>% 
                            arrange(Var1,Var2),1,function(X) paste0(X,collapse="")))
print(cvars)

# combine the df into a big df
df <- df <- dplyr::bind_rows(c(p1))
```

# Generate spectral profiles

```{r}
# check the df structure is fine
# str(df)
# summary(df$medians.max.medians.)
# table(df$sample)
# table(df$`rownames(df)`)
# table(df$medians.max.medians.)

# clean up the sample names
df$sample <- gsub(" ", "_", df$sample)

markers <- unique(df$sample)
```

### Add AF in 

Import AF profile

```{r}
af <- readRDS("./rds/unstained_lympho_spectrum.rds")
af$sample <- "AF"
# Add AF to the spectral profile data frame
df <- rbind(df, af)
```


### Calculate the unmixing matrix

Construct the spectral matrix M.

```{r}
df <- df %>% pivot_wider(names_from = sample, values_from = medians.max.medians.)
M <- as.matrix(df[,2:23])
```

## Construct the weight matrix by variance metrics, 1/CV(channels), and use the standard solution 20260818

```{r}
# weights
cvs <- rowSds(M)/abs(rowMeans(M))
Weights <- 1/cvs
W <- diag(Weights)
dim(W)

# Calculate M^T
M <- as.data.table(M)
M_t <- transpose(M)
M <- as.matrix(M)
M_t <- as.matrix(M_t)
rownames(M_t)<- colnames(M)

# weighted solution
# the weighted M inverse matrix
wmtm_1 <- solve(M_t %*% W %*% M)


# Transpose to make the cells on rows again to match the original data
t_unmixed_matrix <- t(unmixed_matrix)
colnames(t_unmixed_matrix) <- gsub("_","_unmixed_", colnames(t_unmixed_matrix))

#Add these columns to the Time and FSC, SSC channels
unmixed_data <- cbind(exprs(sample[,sample@parameters@data$name[!sample@parameters@data$name %in% df$`rownames(df)`]]),
      t_unmixed_matrix)
```


Load in raw sample data.

```{r}
fs_path <- "D:/YL17_R unmixied/Colon500 UM CD14-_Full-stain_2025_12_23_14_24_41.fcs"
sample <- read.FCS(fs_path, transformation = "linearize-with-PnG-scaling")
```

Obviously, 'Time',  "SSC-H"   "SSC-A"   "FSC-H"   "FSC-A" "SSC-B-H" "SSC-B-A" are not included in constructing the unmixing matrix.

```{r}
# take only the fluorescence channels to go into the unmixing
sample_fluo<-exprs(sample[,sample@parameters@data$name[sample@parameters@data$name %in% aurora_channels]])
as.matrix(sample_fluo)

dim(sample_fluo)

colnames(sample_fluo) <- NULL

# transpose it into the cells as columns vs channels as rows format
sample_fluo<- as.data.table(sample_fluo)
sample_fluo<- transpose(sample_fluo)
sample_fluo<-as.matrix(sample_fluo)

# Solution
unmixed_matrix <- wmtm_1 %*% M_t %*% W %*% sample_fluo

# Transpose to make the cells on rows again to match the original data
unmixed_matrix <- as.data.table(unmixed_matrix)
unmixed_matrix <- transpose(unmixed_matrix)
t_unmixed_matrix <- as.matrix(unmixed_matrix)
rm(unmixed_matrix, sample_fluo)

colnames(t_unmixed_matrix) <- gsub("_","_unmixed_", rownames(wmtm_1))

#Add these columns to the Time and FSC, SSC channels
#exprs(sample[,sample@parameters@data$name[!sample@parameters@data$name %in% df$`rownames(df)`]])
unmixed_data <- cbind(exprs(sample[,sample@parameters@data$name[!sample@parameters@data$name %in% aurora_channels]]),
      t_unmixed_matrix)
rm(p1,dat)
```
Save the unmixing results.

```{r}
# write the fcs file
ff <- flowFrame(unmixed_data)
sample_name <- "CLP500"
write.FCS(ff, filename = paste0("20260818_wls_unmixed_byCV_", sample_name, ".fcs"))

# Save the unmixing results
wls <- list()
wls$CV_weights <- W
wls$CV_weighted_covariance_inverse <- wmtm_1
wls$unmixed_data_weighted_byCV <- unmixed_data
  
saveRDS(wls,paste0("./rds/20260818_WLS_unmixed_",sample_name, ".rds"))

```


## Construct the weight matrix by variance metrics, 1/CV(channel), 20260427

Load the previous data back in.

```{r}
wls <- readRDS("./rds/20260323_WLS_unmixed_CUM500.rds")
M <- wls$Spectral_matrix
```

Calculate the Weights by 1/CV(M) then calculated Weighted M.

```{r}
cvs <- rowSds(M)/rowMeans(M)
Weights <- 1/cvs
W <- diag(Weights)

# Apply weights
WM <- W %*% M
```
Calculate pseudo-inverse of M

```{r}
# SVD for weighted M and calculating pseudoinverse of the weighted M
svd_WM <- svd(WM)
pseudo_inver_WM <- (svd_WM$v) %*% diag(svd_WM$d^(-1)) %*% t(svd_WM$u)
rownames(pseudo_inver_WM) <- colnames(wls$Spectral_matrix)

```

Visualise the unmixing matrix.

```{r}
sample_name <- "CUM500"
aurora_channels <- readRDS("./rds/aurora_channels.rds")
colnames(pseudo_inver_WM)<- aurora_channels

pheatmap(pseudo_inver_WM, 
         main = "Unmixing matrix by WLS",
         color = colorRampPalette(c("blue", "white", "red"))(100),
         cluster_rows = FALSE,
         cluster_cols = FALSE,
         angle_col = c("45"),
         filename = paste0("./plots/ChcvWLS_unmixing_matrix_", sample_name, ".png"),
       #  cellwidth = 5.5,
        width = 12,
        height = 4)
```


Load in raw sample data.

```{r}
fs_path <- "Colon500 UM CD14-_Full-stain_2025_12_23_14_24_41.fcs"
sample <- read.FCS(fs_path, transformation = "linearize-with-PnG-scaling")
```

```{r}
# take only the fluorescence channels'values to go into the unmixing
sample_fluo<-exprs(sample[,sample@parameters@data$name[sample@parameters@data$name %in% aurora_channels]])
dim(sample_fluo)
colnames(sample_fluo) <- NULL

# transpose it into the cells as columns vs channels as rows format
sample_fluo<- as.data.table(sample_fluo)
sample_fluo<- transpose(sample_fluo)
sample_fluo<-as.matrix(sample_fluo)

# Applying unmixing matrix
unmixed_matrix <- pseudo_inver_WM %*% sample_fluo

# Transpose to make the cells on rows again to match the original data
unmixed_matrix <- as.data.table(unmixed_matrix)
unmixed_matrix <- transpose(unmixed_matrix)
t_unmixed_matrix <- as.matrix(unmixed_matrix)
rm(unmixed_matrix, sample_fluo)

colnames(t_unmixed_matrix) <- gsub("_","_unmixed_", rownames(pseudo_inver_WM))

#Add these columns to the Time and FSC, SSC channels
#exprs(sample[,sample@parameters@data$name[!sample@parameters@data$name %in% df$`rownames(df)`]])
unmixed_data <- cbind(exprs(sample[,sample@parameters@data$name[!sample@parameters@data$name %in% aurora_channels]]),
      t_unmixed_matrix)
```
Save the unmixing results.

```{r}
# write the fcs file
ff <- flowFrame(unmixed_data)
write.FCS(ff, filename = paste0("2026427_wls_unmixed_bychannelCV_", sample_name, ".fcs"))

# Save the unmixing results
wls$ChnlCV_weights <- W
wls$ChnlCV_weighted_spectral_matrix <- WM
wls$weighted_unmixing_matrix_bychanlCV <- pseudo_inver_WM
wls$unmixed_data_weighted_bychanlCV <- unmixed_data
  
saveRDS(wls,paste0("./rds/20260427_WLS_unmixed_",sample_name, ".rds"))

```
